Fast Sorting on Modern Computers

### 2005-08-26 docs/tech_report/conclusions.tex
most of davids edits in

However, his move to multi-quicksort, while not affecting the rate of branch
mispredictions in the partitioning phase, results in a large increase in branch
mispredictions due to the addition of a binary search. We determine that this
while reducing the instruction count, the increase is actual running time due to
branch mispredictions is significant.

@@

LaMarca's initial improvements to base mergesort were due to reduce the
instruction count. We noted a reduction with each of the steps he prescribed. We
note too that each of his improvement reduce level 2 cache misses, and the shape
of our level 2 cache miss graphs exactly match his. We also note similar
characteristics: LaMarca mentions a wobble in the instruction count graphs, due to
having to copy the final merge back; we note this same wobble.

>>>>

However, LaMarca's move to multi-quicksort, while not affecting the rate of
branch mispredictions in the partitioning phase, results in a large increase in
branch mispredictions due to the addition of a binary search. We determine that
this while reducing the instruction count, the increase is actual running time
due to branch mispredictions is significant.

@@

LaMarca's initial improvements to base mergesort were expected to reduce the
instruction count. We noted a reduction with each of the steps he prescribed. We
note too that each of his improvement reduce level 2 cache misses, and the shape
of our level 2 cache miss graphs exactly match his. We also note similar
characteristics: LaMarca mentions a wobble in the instruction count graphs, due to
having to copy the final merge back; we note this same wobble.

### 2005-08-26 docs/tech_report/shell.tex
most of davids edits in

Shellsort is a derivative of insertion sort designed to offer a compromise
between the performance of mergesort and radixsort, and the space efficiency of
selection sort, bubblesort and insertion sort. It was designed by Donald Shell
in 1959, and described in \cite{Shell59}\footnote{Shell describes the former
sorts as \n{Merging Pairs} and \n{Floating Decimal Sort}, respectively, and the
latter sorts as \n{Finding the Smallest}, \n{Interchanging Pairs} and
\n{Sifting}, respectively.}.

@@

Since first published, there has been much discussion about the best
\n{increments} to use in shellsort. Shell's original sort, described above,
has the increments $\frac{N}{2}$, $\frac{N}{4}$, $\frac{N}{8}$, $\frac{N}{16}$
........ $1$. \cite{Sedgewick96} discusses many increments used, and provides
code for comparing these. Based on running the \cc{driver} program\footnote{See
http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
seen to be the fastest on the authors local machine. As a result, this was used
to test the performance of shellsort. This uses increments which are
$\frac{5}{11}^{th}s$ of the previous increment.

@@

The cache performance of shellsort should be regular. Each sub-array is sorted in
turn without reference to the next. Since each element of a sub-array is quite
far from its next element, an entire cache line must be loaded for each key,
which will be used once per iteration of the sort across the sub-array. If there
are a large number of sub-arrays, as during the iterations with large increments,
there will be no spatial reuse, and little temporal reuse, so long as the
sub-array is large than the cache. Once the increment is small enough that each
sub-array fits within the cache (i.e. quickly for L2 cache, and much more slowly
for the L1 cache), then temporal reuse will occur. Once an entire sub-array fits
within an eighth of the size of the cache (eight being the number of keys
that fit in a cache line), then spatial reuse will occur as well. This reuse
will only occur within a single iteration; once the increment with the next
increment begins, most of the contents of the cache will be invalidated, unless
the entire array fits within the cache.

@@

Improved shellsort does improve on shellsort, having approximately 10\% fewer
instructions. It also has slightly fewer cache misses and branch mispredictions.
Overall, this leads to a reduction in cycle count of between 35 and 70\%, as
seen in Figure \ref{shell cycles}.

>>>>

Shellsort is a derivative of insertion sort designed to offer a compromise
between the performance of mergesort and radixsort\footnote{Quicksort and
heapsort were developed only later, so at the time there was no efficient
in-place sort.}, and the space efficiency of selection sort, bubblesort and
insertion sort. It was designed by Donald Shell in 1959, and described in
\cite{Shell59}\footnote{Shell describes the former sorts as \n{Merging Pairs}
and \n{Floating Decimal Sort}, respectively, and the latter sorts as \n{Finding
the Smallest}, \n{Interchanging Pairs} and \n{Sifting}, respectively.}.

@@

Since first published, there has been much discussion about the best
\n{increments} to use in shellsort. Shell's original sort, described above,
has the increments $\frac{N}{2}$, $\frac{N}{4}$, $\frac{N}{8}$, $\frac{N}{16}$
........ $1$. \cite{Sedgewick96} discusses many increments used, and provides
code for comparing these. Based on running the \cc{driver} program\footnote{See
http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
seen to be the fastest on the authors' local machine. As a result, this was used
to test the performance of shellsort. This uses increments which are
$\frac{5}{11}^{th}s$ of the previous increment.

@@

The cache performance of shellsort should be regular. Each sub-array is sorted in
turn without reference to the next. Since each element of a sub-array is quite
far from its next element, an entire cache line must be loaded for each key,
which will be used once per iteration of the sort across the sub-array. If there
are a large number of sub-arrays, as during the iterations with large increments,
there will be no spatial reuse, and little temporal reuse, so long as the
sub-array is larger than the cache. Once the increment is small enough that each
sub-array fits within the cache (i.e. quickly for L2 cache, and much more slowly
for the L1 cache), then temporal reuse will occur. Once an entire sub-array fits
within an eighth of the size of the cache (eight being the number of keys that
fit in a cache line), then spatial reuse will occur as well. This reuse will
only occur within a single iteration; once the increment with the next increment
begins, most of the contents of the cache will be invalidated, unless the entire
array fits within the cache.

@@

Improved shellsort does improve on shellsort, having approximately 10\% fewer
instructions. It also has slightly fewer cache misses and branch mispredictions.
Overall, this leads to a reduction in cycle count of between 35\% and 70\%, as
seen in Figure \ref{shell cycles}.
